Visualisation of Brain Statistics with R-package `ggseg`

### 2019-04-12 draft_1_DVP.Rmd
Update draft_1_DVP.Rmd

```{r void, fig.width=5, fig.cap="ggseg plots can be used with most standard scales, themes and such that work with ggplot. The special ggseg options for hemispheres, view etc. depend on the atlas used, and some options are only available for certain atlases. There is no 'lateral' or 'medial' views of subcortical atlases. "}
p1 <- ggseg(position = "stacked") + 
  theme_dark() + 
  labs(title="dkt", subtitle = "dark theme")
p2 <- ggseg(position = "stacked") + 
  theme_classic() + 
  labs(title="dkt", subtitle = "classic theme")

@@

One have several options for plotting the main brain representations.
These options are atlas-specific. 
For cortical atlases, such as the DKT, one can stack the hemispheres, plot one or both hemisphere, or view only the medial or lateral side (or any combination of these; see Figure \@ref(fig:stacked)).
For subcortical atlases, such as the aseg, the options are more limited but one can often choose between axial, sagital, and coronal views.

```{r stacked, fig.width=5, fig.cap="There are several ggseg-special options that may be supplied to control how the plots looks and is organised. Top left: plot using medial view. Bottom left: plot using left hemisphere. Right: plot combining both the options. "}
lat <- ggseg(view = "medial") + 
  labs(title = "Medial view")
left <- ggseg(hemisphere = "left") + 
  labs(title = "Left hemi.")
pp <- cowplot::plot_grid(lat, left, nrow=2)

@@

### Using own data with fill and colour
`ggseg` accepts any argument you can supply to `geom_polygon` and therefore is easy to work with for those familiar with ggplot functionality. 
Standard arguments like `fill` to flood the segments with a colour, or `colour` to colour the edges around the segments are typical arguments to provide to the function either as a single value or within the ggplot mapping option `aes`.

```{r fill, fig.width=5, fig.height=8, fig.cap="Supplying 'area' to the fill option in ggseg, will use the column 'area' from the accompanying dataset to create a discrete colour palette over the segments in the atlas."}
ggseg(mapping=aes(fill = area), colour="black") + 
  scale_fill_brain("dkt") +
  theme(legend.justification=c(1,0),
        legend.position="bottom",
        legend.text = element_text(size = 5)) +
  guides(fill = guide_legend(ncol = 3))
```

Typically you would want each area to have a different colour, corresponding to some descriptive or inference statistic that has been computed.
The bulk of the package originates not in its functions, but in the datasets that accompany the package. 
There is one dataset per atlas, which contains key information regarding the atlas, and coordinates for the segment polygons of the atlas. 
Before setting up the data from the statistics to project onto the segments, looking at the atlas datasets might help understand what the data needs to look like, or avoid looking like.

@@

The column 'label' was added to the atlas for easy matching with the output from Freesurfer statistics tables. 
The data is in long-format, meaning each segment has its own row, meaning the data you wish to project onto the segments also need to be in long-format. 
As data directly from Freesurfer comes in wide-format, this requires some extra wranling.

@@

This data can now be used directly with the `ggseg` function, as the `label` column corresponds in name and content with the `label` column in the atlas data of dkt.
The data **must** include a column that has the same name at atleast _some_ data matching the values in the corresponding column in the atlas data.
In the next example we create some data with 4 rows, and an 'area' and 'p' column. 
The `ggseg` function will recognise the matching column 'area', and merge the supplied data into the atlas using `dplyr` joins. 
We use the `p` column as the column flooding the segment with colour.

@@

We can also change how this looks, by various ggplot functions, like scales, labs, themes and such. 
```{r datasupp2, fig.width=5}
p +
  theme_void() +
  scale_fill_gradient(low="firebrick",high="goldenrod") +
  labs(title="A nice plot title", fill="p-value")
```

@@

### Creating subplots
If your data has different statistics for different groups, and you want to plot them all, we can use ggplot's `facet_wrap` or `facet_grid` for that.
The data needs to be in long format for this to work, as needed by ggplot, meaning the data from the groups should appear in seperate rows in the data, not in separate columns. Secondly, the data needs to be grouped using `dplyr`'s `group_by` function _before_ providing the data to the `ggseg` function. 
The ggseg-function will detect if the data is grouped, and perform the necessary steps for the `facet`'s to work when they are applied later.

```{r datasupp3, fig.width=5}
someData = data.frame(
  area = rep(c("transverse temporal", "insula",
               "pre central","superior parietal"),2), 
  p = sample(seq(0,.5,.001), 8),
  AgeG = c(rep("Young",4), rep("Old",4)),
  stringsAsFactors = FALSE) %>% 
  group_by(AgeG)

@@

Depending on what type of column is used for fill (continuous or discrete),
you can also use the differenct `scales` to alter the colours.

@@

All the consepts described above also work with the `aseg` atlas for subcrotical structures, except arguments to see only one hemisphere or lateral or medial surfaces, as these arguments do not make sense for this atlas. 
Again, inspecting the atlas data directly will be of aid when preparing data to use with the atlas
```{r , size = 'small', out.width='90%'}
aseg
```

@@

Creating plots of the brain as 2-dimensional can at times make it difficult to properly understand where an atlas segment is located in the true 3-dimensional brain. 
This is particularly true when looking within the brain as subcortical structures or white matter tracts.
To provide the option of viewing and printing the plots in 3-dimensional space, another function was created using [plotly](https://plotly-book.cpsievert.me/).
The data is somewhat more complex than the 2D ggplot polygon version `ggseg`, as it includes options for brain inflation, glass brains and much more, to ease viewing and understanding.
Another positive effect of basing this function on plotly, is the interactivity of the plot, which also makes it nice for public dissemination.
A lot of credit goes to A.M.Winkler and his [Brainder](https://brainder.org/) work, which supplied us with the first examples of going from Freesurfer `.srf`-files to `.ply`-files, and whose scripts massively aided us in making this work.

The function `ggseg3d()`, is based in the plotly, it is recommended to get a little familiarized with with [plotly](https://plotly-book.cpsievert.me/).

>>>>

```{r void, fig.width=5, fig.cap="ggseg plots can be used together with most standard scales, themes and such that work with ggplot. The additional ggseg options for hemispheres, view etc. are atlas-specific. For example 'lateral' or 'medial' views are only available for cortical atlases such as the DKT. "}
p1 <- ggseg(position = "stacked") + 
  theme_dark() + 
  labs(title="dkt", subtitle = "dark theme")
p2 <- ggseg(position = "stacked") + 
  theme_classic() + 
  labs(title="dkt", subtitle = "classic theme")

@@

One has several options for plotting the main brain representations.
These options are atlas-specific. 
For cortical atlases, such as the DKT, one can stack the hemispheres, plot one or both hemisphere, or view only the medial or lateral side (or any combination of these; see Figure \@ref(fig:stacked)).
For subcortical atlases, such as the aseg, the options are more limited but one can often choose between axial, sagital, and coronal views.

```{r stacked, fig.width=5, fig.cap="One can control the organization and the global visualization of a ggseg plot with gggseg-specific options such as hemisphere or view. Top left: plot using medial view. Bottom left: plot using left hemisphere. Right: plot combining both the options. "}
lat <- ggseg(view = "medial") + 
  labs(title = "Medial view")
left <- ggseg(hemisphere = "left") + 
  labs(title = "Left hemi.")
pp <- cowplot::plot_grid(lat, left, nrow=2)

@@

### Using own data with fill and colour
`ggseg` accepts any argument you can supply to `geom_polygon` and therefore is easy to work with for those familiar with ggplot functionality. 
Standard arguments like `fill` to flood the segments with a colour, or `colour` to colour the edges around the segments are typical arguments to provide to the function either as a single value or within the ggplot mapping option `aes`.
To use color palettes corresponding to those used in the original neuroimaging softwares one can use atlas-specific 'brain' palette scales. See an example in Figure \@ref(fig:fill).

```{r fill, fig.width=5, fig.height=8, fig.cap="Supplying 'area' to the fill option in ggseg, will use the column 'area' from the accompanying dataset to create a discrete colour palette over the segments in the atlas. The scale_fill_brain("dkt") option will colour the plot according the original colors used in FreeSurfer."}
ggseg(mapping=aes(fill = area), colour="black") + 
  scale_fill_brain("dkt") +
  theme(legend.justification=c(1,0),
        legend.position="bottom",
        legend.text = element_text(size = 5)) +
  guides(fill = guide_legend(ncol = 3))
```

[comment]: <> (Starting here I do not comment, as maybe re-structuring is better; waiting for input)
Most users will use ggseg to display through a color scale some descriptive or inferential statistics such as mean cortical thickness or thickness – cognition relationship across the different brain regions.
The bulk of the package originates not in its functions, but in the datasets that accompany the package. 
There is one dataset per atlas, which contains key information regarding the atlas, and coordinates for the segment polygons of the atlas. 
Before setting up the data from the statistics to project onto the segments, looking at the atlas datasets might help understand what the data needs to look like, or avoid looking like.

@@

[comment]: <> (End of no comments, as maybe re-structuring is better; waiting for input)
In any atlas, the column ‘label’ is particularly useful for combining the data of interest with the ‘ggseg’ polygons. 
The column ‘label’ contains the label (region) names as in the original neuroimaging software. 
For example, the DKT atlas label column matches the region names from Freesurfer statistics tables.
Yet the data in ‘ggseg’ is in a long format - that is each region has its own row - thus data of interest needs to be converted into this format. 
Note that most datasets are organized in wide format, in which subjects are represented by rows and each different data variable is represented in a separate column. 
See below an example of wide-to-long conversion.

@@

This data - in long format - can now be used directly with the `ggseg` function, as the `label` column corresponds in name and content with the `label` column in the atlas data of dkt.
The data **must** include a column that has the same name at least _some_ data matching the values in the corresponding column in the atlas data.
In the next example we create some data with 4 rows, and an 'area' and 'p' column, representing the results of an analysis. 
The `ggseg` function will recognise the matching column 'area', and merge the supplied data into the atlas using `dplyr` joins. 
We use the `p` column as the column flooding the segment with colour.

@@

We can then modify the appearance similarly to any other ggplot graph using ggplot functions such as scales, labs, themes, etc. 
```{r datasupp2, fig.width=5}
p +
  theme_void() +
  scale_fill_gradient(low="firebrick",high="goldenrod") +
  labs(title="A nice plot title", fill="p-value")
```

@@

### Creating subplots
One can plot a statistic of interest in different groups (i.e. thickness or significance in young or older adults) using ggplot's `facet_wrap` or `facet_grid`.
As before, data needs to be in long format (group data should appear in seperate rows, not in separate columns). 
Secondly, the data needs to be grouped using `dplyr`'s `group_by` function _before_ providing the data to the `ggseg` function. 
The ggseg function will detect grouped data, and adapt it to `facet`'s requirements.

```{r datasupp3, fig.width=5}
someData = data.frame(
  area = rep(c("transverse temporal", "insula",
               "pre central","superior parietal"),2), 
  p = sample(seq(0,.5,.001), 8),
  AgeG = c(rep("Young",4), rep("Older",4)),
  stringsAsFactors = FALSE) %>% 
  group_by(AgeG)

@@

As before, one can apply the different ggplot and facet options, such as `scales` to modify the plot's appearance.

@@

All the concepts described above also work with the `aseg` atlas for subcortical structures, except hemisphere and view arguments that do not apply in subcortical atlases such as aseg. 
Again, inspecting the atlas data directly will be of aid when preparing data to use with the atlas
```{r , size = 'small', out.width='90%'}
aseg
```

@@

Representing brains as 2D polygons is a good solution for fast, efficient and flexible plotting, based on ggplot (a powerful tool for visualization) and that can be easily combined with interactive apps such as in shiny.
Yet brains are intrinsically 3D and it can be challenging to recognize the location of a region. This problem exacerbated in atlases representing subcortical structures – after all, cortical surfaces are 2D – such as grey matter structures or white matter tracts. 
Hence, here we also provide the ggseg3D function to plot, view and print 3D atlases in R. 
ggseg3D is based on tri-surface mesh plots using plotly. 
The data structure is somewhat more complex than the 2D ggplot polygon version and includes additional options for brain inflation, glass brains, camera locations, etc.  
As ggseg3D is based on plotly, the resulting brain atlases are interactive, which guides interpretation, and is useful for public dissemination. 
We recommend users to get a little familiarized with with [plotly](https://plotly-book.cpsievert.me/).
